Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile

### 2025-05-27 advances/informe2.tex
up inf2

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
  pdfauthor={Andreas Laffert},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Preferences for market justice: the role of intergenerational
social mobility and beliefs in meritocracy in Chile}

  \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }

@@

What makes market-driven inequalities appear legitimate in the eyes of
the public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that access to
services like health care, pensions, and education be governed by
market-based criteria---a phenomenon known as \emph{market justice
preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing literature shows that market justice preferences are shaped by
both the economic and institutional context and individuals' positions
within social stratification. Grounded in the notion that institutions
influence individual attitudes
(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
find that countries with stronger public provision or more expansive
welfare states exhibit lower levels of market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
more privatized contexts show stronger support for market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}).

@@

Although it is clear that one's social position influences market
preferences, the question of how upward or downward mobility within the
social structure affects these preferences remains unanswered. This
question is far from trivial, especially in Latin America, where many
have experienced various forms of mobility amid high economic inequality
and deep welfare privatization (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}; \citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \&
Fachelli, 2021}; \citeproc{ref-torche_intergenerational_2014}{Torche,
2014}). Social origins and destinations affect attitudes toward
inequality in distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske,
2017}; \citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
2016b}, \citeproc{ref-gugushvili_subjective_2017}{2017};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
movement between these positions exposes individuals to different
experiences and mechanisms that shape their views on what is fair
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
research, examining the effects of social mobility on market justice
preferences can help to illuminate how inequalities in access to social
services are justified among individuals who have experienced, or not,
changes in their social standing, and what are the normative mechanisms
that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
2015}).

Beyond their isolated effects, social mobility and meritocratic beliefs
interact in complex ways to shape market justice preferences. A key
mechanism proposed in the literature to explain how mobility influences
distributive justice preferences is the psychological process of
self-serving bias in causal attribution
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
suggests that individuals attribute failures---such as downward
mobility---to external factors, while crediting successes---such as
upward mobility---to their own merit and effort
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
who experience upward mobility tend to view their social position as
earned, making them more likely to believe that individuals are
responsible for their own success or failure. Research shows that upward
mobility is associated with weaker preferences for redistribution
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
2011}) and stronger legitimacy of income inequality
(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
individuals who experience downward mobility tend to blame structural
factors like inequality and are more supportive of redistribution while
rejecting merit-based explanations
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
together, I argue that meritocratic beliefs may reflect---or
reinforce---this self-serving attribution mechanism by legitimizing
one's social status as the outcome of personal merit, closely tied to
attribution bias.

@@

This study focuses on Chile, a particularly intriguing case for
examining market justice preferences. Despite significant economic
growth and poverty reduction, Chile has some of the highest levels of
inequality in Latin America and among OECD countries
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
coexists with short-range upward mobility among lower-class segments
moving into middle strata, though strong barriers remain to reaching
higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
2014}). What makes Chile especially salient is that much of this
inequality is rooted in deep neoliberal reforms that institutionalized
the privatization and commodification of key social sectors
(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
during the dictatorship (1973--1989) and expanded in democracy, these
reforms enabled the unprecedented emergence of markets in health,
pensions, and education, with provision segmented by individuals'
ability to pay and supported by public subsidies
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
despite waves of protest against inequality and commodification from
2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
2021})---Chilean subjectivities have been increasingly shaped by
neoliberal discourses and market logics, influencing their attitudes
toward inequality and welfare distribution
(\citeproc{ref-arteagaaguirre_politicas_2015}{Arteaga Aguirre \& Iñigo
Valderrama, 2015}; \citeproc{ref-castillo_socialization_2024}{Castillo
et al., 2024}).

@@

@@

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
  pdfauthor={Andreas Laffert Tamayo},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Preferences for mercantilization: the role of intergenerational
social mobility and beliefs in meritocracy in Chile}

  \author{Andreas Laffert Tamayo}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }

@@

What is the legitimate extent of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that the
allocation of services like health care, pensions, and education be
governed by market-based criteria---a phenomenon known as \emph{market
justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing literature shows that market justice preferences are shaped by
both the economic and institutional context and individuals' positions
within social stratification. Grounded in the notion that economic
institutions influence people's normative attitudes
(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
find that countries with stronger public provision or more expansive
welfare states exhibit lower levels of market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
more privatized contexts show stronger support for market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}).

@@

Although it is clear that one's social position influences market
justice preferences, the question of how upward or downward mobility
within the social structure affects these preferences remains
unanswered. This question is far from trivial, especially in Latin
America, where many have experienced various forms of mobility amid high
economic inequality and deep welfare privatization
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}).
Social origins and destinations affect attitudes toward inequality in
distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b},
\citeproc{ref-gugushvili_subjective_2017}{2017};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
movement between these positions exposes individuals to different
experiences and mechanisms that shape their views on what is fair
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
research, examining the effects of social mobility on market justice
preferences can help to illuminate how inequalities in access to social
services are justified among individuals who have experienced, or not,
changes in their social standing, and what are the normative mechanisms
that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
2015}).

Beyond their isolated effects, social mobility and meritocratic beliefs
interact in complex ways to shape market justice preferences. Among
others, a key mechanism proposed in the literature to explain how
mobility influences distributive justice preferences is the
psychological process of self-serving bias in causal attribution
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
suggests that individuals attribute failures---such as downward
mobility---to external factors, while crediting successes---such as
upward mobility---to their own merit and effort
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
who experience upward mobility tend to view their social position as
earned, making them more likely to believe that individuals are
responsible for their own success or failure. Research shows that upward
mobility is associated with weaker preferences for redistribution
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
2011}) and stronger legitimacy of income inequality
(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
individuals who experience downward mobility tend to blame structural
factors like inequality and are more supportive of redistribution while
rejecting merit-based explanations
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
together, I argue that meritocratic beliefs may reinforce this
self-serving attribution mechanism by legitimizing one's social status
as the outcome of personal merit, closely tied to attribution bias.

@@

This study focuses on Chile, a particularly intriguing case for
examining market justice preferences. Despite significant economic
growth and poverty reduction, Chile has some of the highest levels of
inequality in Latin America and among OECD countries
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
coexists with short-range upward mobility among lower-class segments
moving into middle strata, though strong barriers remain to reaching
higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
2014}). What makes Chile especially salient is that much of this
inequality is rooted in deep neoliberal reforms that institutionalized
the privatization and commodification of key social sectors
(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
during the dictatorship (1973--1989) and expanded in democracy, these
reforms enabled the unprecedented emergence of markets in health,
pensions, and education, with provision segmented by individuals'
ability to pay and supported by public subsidies
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
despite waves of protest against inequality and commodification from
2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
2021})---Chilean subjectivities have been increasingly shaped by
neoliberal discourses and market logics, influencing their attitudes
toward inequality and welfare distribution
(\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli, 2012};
\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

@@

Although redistribution typically refers to the state's capacity to
reallocate resources from the advantaged to the vulnerable, market
institutions also play a central role in distributing socially valuable
goods and rewards (Koos \& Sachweh, 2019; Lindh \& McCall, 2023). As
Polanyi (1957) observed, economic integration in capitalist societies is
primarily organized through market exchange, governed by a
self-regulating price system embedded in institutional frameworks. These
institutions are not mere aggregates of individual behavior, but social
realities endowed with rules, mechanisms, and normative meanings that
shape everyday thinking (Koos \& Sachweh, 2019; Immergut, 1998). In this
sense, the economic order is mirrored in a moral economy: collectively
shared norms and beliefs about justice in distribution, embedded and
reinforced through institutions (Mau, 2015; Svallfors, 2006). While most
research from this perspective has focused on welfare institutions,
recent scholarship has brought the market back into focus as a site of
distributive justice beliefs and institutional responses to inequality
(Lindh \& McCall, 2023). In many countries, privatization and
commodification have expanded market logic into core areas of social
reproduction, such as healthcare, education, pensions, and social
security, deepening inequality in access to these services (Gingrich,
2011; Ferre, 2023). Yet public support for market-based welfare
provision has grown, even in traditional welfare states, particularly
among higher-income groups who view private alternatives as more
efficient or higher in quality (Busemeyer \& Iversen, 2020). This shift
calls for broader inquiry into how institutions like markets structure
access to resources and legitimize inequalities (Lindh \& McCall, 2023;
Mau, 2015).

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also examines the beliefs individuals hold about the causes
of social inequality, the normative principles that underpin those
beliefs, the factors that shape them, and their consequences for
attitudes and behavior (Kluegel \& Smith, 1981). In this regard,
empirical research on distributive justice focuses on individuals'
conceptions of how goods and rewards \emph{should} be distributed in
society (Jasso et al., año). This line of inquiry allows for examining
the extent to which economic inequality is perceived as just, or in a
certain sense, legitimate (Castillo, 2011). One prominent strand
investigates the legitimacy of wage inequality, particularly salary
differences across occupations (Jasso, 1978; Castillo, 2011; Wegener,
1999). Another examines how individuals justify market-generated
inequalities in access to core social services such as healthcare,
education, and pensions. Here, legitimacy stems from the belief that
access to these goods should follow market-based criteria (Castillo et
al., 2024; Lindh, 2015). In such views, these services are treated as
legitimate commodities; goods that can be traded, priced, and evaluated
through market logic (Busemeyer \& Iversen, 2020).

Market justice preferences refer to normative beliefs that legitimize
the idea that access to core social services---such as healthcare,
education, or pensions---should be determined by market-based criteria.
Following Janmaat's (2013, p.~359) distinction, these preferences fall
under the category of ``beliefs,'' understood as normative ideas about
what inequality should look like, as opposed to ``perceptions,'' which
refer to subjective evaluations of existing inequality. Market justice
preferences reflect the view that access to these services should depend
on individuals' ability to pay, thus justifying inequalities generated
by market mechanisms (Kluegel \& Smith, 1999; Lindh, 2015). The concept
draws on Lane's (1968) classic contrast between market justice and
political justice: the former is grounded in the idea of earned
deserts---where rewards reflect effort, productivity, and skill---while
the latter prioritizes need and equality, typically expressed in welfare
state policies. Lane argued that markets and states differ in purpose
(efficiency vs.~need), logic (individual vs.~collective), and fairness
criteria (merit vs.~equality). Market justice assumes that markets are
neutral, self-regulating systems in which fair procedures yield outcomes
proportional to merit. Inequality, from this perspective, is not only
expected but legitimate---as long as it arises from fair competition. In
this way, market justice offers a moral lens through which individuals
can view the commodification and stratified access to social services as
fair and justified (Kluegel \& Smith, 1999; Lindh, 2015).

The empirical study of market justice preferences has relied on diverse
strategies to operationalize how individuals evaluate inequalities
produced by market allocation. One common approach assesses whether
people consider it fair that access to key social services---such as
healthcare, education, or pensions---depends on income. This strategy
builds on foundational work by Kluegel and Smith (1981), who explored
the normative justifications behind support for economic inequality.
More recent studies have expanded this logic beyond income to include
market-mediated access to welfare goods. For instance, Immergut and
Schneider (2020) and von dem Knesebeck et al.~(2016) examine whether
respondents believe it is just that those with higher incomes receive
better healthcare. Similarly, Lee and Stacey (2023) apply this approach
to educational opportunities. These studies typically use targeted
survey items that ask respondents to evaluate specific distributive
scenarios, allowing for comparative analyses across countries or welfare
regimes. Complementing these efforts, more recent research has
introduced composite indicators designed to capture broader orientations
toward market-based allocation. For example, Castillo et al.~(2024)
propose a single-item measure that summarizes individual support for
income-based access across multiple domains---health, education, and
pensions---within the Chilean context. Together, these instruments aim
to capture to what extent individuals perceive market-generated
inequality as justified.

Comparative empirical research has identified several individual-level
factors that influence support for market justice. Individuals in more
advantaged socioeconomic positions---those with higher income,
education, and occupational status---are consistently more likely to
endorse market-based distributive principles (Svallfors, 2007; Koos \&
Sachweh, 2019; Lindh, 2015). For example, Lindh (2015) finds that
individuals from the service class are more likely to support
market-based access to healthcare and education than skilled and
unskilled workers across 17 relatively affluent countries. In a
comparative analysis, Svallfors (2007) observes that this expected class
pattern appears clearly only in Sweden, where support for private
education and healthcare varies systematically by class. Busemeyer
(2014) similarly shows that support for private education is stronger
among high-income groups, while Immergut and Schneider (2020) and von
dem Knesebeck et al.~(2016) report comparable findings for healthcare,
suggesting that wealthier individuals perceive private provision as a
means of maintaining relative advantage. In Chile, Otero and Mendoza
(2024) show that individuals with higher income and university education
express stronger support for market allocation in healthcare, education,
and pensions. Beliefs about inequality and political orientation also
matter. In Chile, Castillo et al.~(2024) show that individuals who have
strong meritocratic beliefs are more likely to support market-based
distribution, while Lee and Stacy (2023) in Australia suggest that these
preferences are also greater as people lean toward economic
conservatism. In this sense, market justice preferences are shaped by
the interplay between structural position and normative reasoning.

However, individual characteristics alone do not fully explain variation
in support for market justice. A growing body of cross-national research
highlights the importance of institutional arrangements in shaping these
preferences. For instance, Immergut and Schneider (2020) find that in
countries with higher public spending on healthcare, individuals are
less likely to view income-based access as fair. Similarly, Busemeyer
(2014) shows that increased public investment in education is associated
with lower support for privatized provision. Conversely, Lindh (2015)
finds that in countries with more market-oriented welfare systems,
support for market-based distribution tends to be higher, suggesting
that individual attitudes often align with institutional outputs. These
findings are consistent with neo-institutionalist and policy feedback
theories (Campbell, 2012), which argue that institutions do more than
redistribute resources---they also shape the normative categories
through which individuals assess who is deserving of support (Immergut,
1990). In this view, dominant values and preferences are both shaped by
and embedded in institutional configurations (Busemeyer, 2014),
reinforcing the notion that institutions are not neutral structures, but
active producers of the moral frameworks that legitimize or challenge
inequality.

@@

The study of social mobility, its drivers and consequences has long been
central to sociology. Classic theorists explored not only movement
across social hierarchies but also its broader implications for class
conflict, norm stability, and institutional change (Breen \& Ermisch,
2024). Sorokin (1927) introduced the concept formally, defining mobility
as the shift of individuals, values, or objects between positions within
a stratification system, and distinguishing between horizontal and
vertical forms. Later work differentiated intergenerational from
intragenerational mobility, as well as absolute mobility---driven by
structural change---from relative mobility, which captures the extent to
which origins constrain destinations (Eyles et al., 2022). In Latin
America, research shows high absolute but low relative mobility (Bucca,
2016): although educational expansion and economic modernization have
enabled some upward movement, status reproduction remains strong,
especially among elites (Torche, 2015; López-Roldán \& Fachelli, 2021).
This reflects deep structural inequality, segmented education systems,
stratified labor markets, and legacies of dependent development that
restrict access to mobility channels and reinforce the intergenerational
transmission of advantage.

Beyond mapping mobility patterns, growing research has examined its
subjective and attitudinal effects. Mobility effects---defined as
outcomes resulting from movement between origin and destination classes
(Breen \& Ermisch, 2024)---have long attracted theoretical interest.
Sorokin's (1959) dissociative hypothesis posits that mobility, whether
upward or downward, may produce psychological strain due to conflicting
norms between class contexts, leading to lower life satisfaction.
Similarly, Lenski (1954) argued that status inconsistency---mismatches
among education, income, and occupation---can undermine well-being.
These ideas underpin extensive empirical work linking intergenerational
mobility to outcomes such as life satisfaction, mental health, and
stress (Hadjar, Gugushvili, Präg \& Gugushvili).

Research on intergenerational social mobility has increasingly examined
its effects on attitudes toward economic inequality. A key area of
inquiry focuses on how upward and downward mobility influence support
for redistribution, though findings remain mixed. Alesina et al.~(2018)
show that individuals with pessimistic expectations about their
mobility---particularly those anticipating downward movement---are more
likely to support generous redistributive policies. Similarly, Ares
(2020) finds that upwardly mobile individuals tend to be less supportive
of state-led redistribution compared to those who have experienced
downward mobility. Comparative studies by Schmidt (2011) and Gugushvili
(2016) likewise report that subjective upward mobility is associated
with weaker preferences for redistribution, while downward mobility
strengthens redistributive support. However, recent causal evidence from
Breen and Ermisch (2024) suggests the opposite: upward mobility may
increase redistributive preferences, while downward mobility may reduce
them.

Beyond redistribution, other studies have explored the impact of
mobility on broader beliefs about inequality. Gugushvili (2016) finds
that upwardly mobile individuals are more likely to adopt
individualistic attributions of poverty and to legitimize income
inequality, particularly in post-socialist societies. Similarly, Bucca
(2016) shows that subjective upward mobility reinforces individualistic
explanations of wealth in seven Latin American countries. At the macro
level, Shariff et al.~(2016) demonstrate that higher levels of national
economic mobility correlate with greater tolerance of inequality. In
contrast, Day and Fiske (2019) find that low perceived mobility
undermines belief in meritocracy and a just world, thereby weakening
system justification. Taken together, this body of research suggests
that mobility shapes attitudes toward inequality and justice through
multiple, and sometimes contradictory, mechanisms.

The literature on the effects of social mobility has proposed various
mechanisms to explain how and why changes in social position may
influence individual outcomes (Freyr \& Rhem, año). One of the most
prominent is the self-interest mechanism, which posits that individuals
who experience upward or downward mobility undergo a shift in their
material interests, thereby altering their perceptions and preferences
(Ares, 2020; Langsæther, Evans \& O'Grady, 2021; Freyr \& Rhem, año).
Closely related to this logic is the Prospect of Upward Mobility (POUM)
hypothesis, which suggests that individuals may oppose redistribution
not because of their current position, but because they anticipate
improving their status in the future (REF). A second line of explanation
draws on the framework of theories of socialization. Within this
tradition, hypotheses such as acculturation, socialization, and status
maximization propose that individuals adjust their attitudes based on
the norms and values of either their class of origin or their
destination, or a combination of both (Jaime-Castillo \&
Marqués-Perales, 2019). However, most of these mechanisms focus
primarily on the indirect effects of origin and destination positions,
rather than on the direct effect of experience of movement itself. In
response to this gap, recent research has highlighted the role of
self-serving bias in causal attribution processes, suggesting that
individuals tend to explain their mobility trajectories in ways that
justify their current position, which in turn shapes their beliefs and
preferences (Schmidt, 2011; Gugushvili, 2016; Molina, Bucca \& Macy,
2019).

The self-serving bias mechanism builds on a value-oriented perspective,
emphasizing that individuals' experiences of social mobility shape their
causal attributions, which in turn influence their beliefs about justice
and distributive preferences (Gugushvili, 2014). Causal attribution
refers to the process through which individuals generate explanations
for their own behavior and outcomes, as well as those of others
(Gugushvili, year). In this view, people's interpretations of economic
inequality depend on whether they believe such disparities reflect
unequal individual contributions. Individuals who adopt an internal
attribution framework tend to see success or failure as rooted in
personal characteristics such as effort, talent, or merit. In contrast,
those who regard inequality as unjust are more likely to adopt an
external attribution model, viewing outcomes as the result of structural
barriers beyond individual control (Kluegel \& Smith, 1999). This
mechanism, often described as intrapersonal causal attribution, reflects
how people explain their own socioeconomic positions---typically
attributing their successes to internal qualities while blaming failures
on external circumstances (Miller \& Ross, 1975). Over time, individuals
may revise their beliefs and attitudes: while early views are shaped by
their social origin, these are later adjusted in light of personal
experiences of mobility and the perceived role of ascribed versus
achieved factors in determining socioeconomic outcomes (Gugushvili,
year).

Empirical research has provided support for the self-serving bias
mechanism in shaping redistributive preferences and attitudes toward the
legitimacy of inequality. For instance, Schmidt (2011) finds that
individuals who experience upward mobility are more likely to interpret
their success as the result of personal effort or merit, and
consequently perceive redistribution as less necessary. Conversely,
individuals who experience downward mobility tend to attribute their
decline to external circumstances---such as structural inequality or
unemployment---and show stronger support for redistribution. In a
comparative analysis across different welfare domains, Gugushvili
demonstrates that upward mobility is associated with lower support for
government spending on housing and pensions, while individuals who
experience downward mobility express lower support for healthcare and
education spending, but favor increased investment in housing and
pensions, reflecting the material nature of these domains. Moreover,
Gugushvili (2016) finds that upward mobility is linked to greater
justification of income inequality, suggesting that improved social
standing reinforces an attributional view in which success is seen as
the result of individual characteristics---thus legitimizing inequality
as a fair outcome.

These findings suggest that causal attribution processes mediate the
relationship between mobility experiences and justice-related attitudes.
Accordingly, this study proposes the following hypothesis:

\(H1\): Experiencing upward (downward) social mobility is positively
(negatively) associated with greater support for market justice in
healthcare, pensions, and education.

@@

Meritocracy constitutes a central ideological framework for legitimizing
different types of social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1962}{\textbf{young\_rise\_1962?}}), the term
was meant to critique a system in which merit-based stratification
becomes a new form of inequality. However, over time, meritocracy has
been widely supported in many societies as a fair and desirable
principle of distribution, particularly within liberal democracies and
market-oriented economies (\citeproc{ref-sandel_tyranny_2020}{Sandel,
2020}; \citeproc{ref-mijs_paradox_2021}{\textbf{mijs\_paradox\_2021?}}).
From a sociological perspective, the belief in meritocracy is more than
a cognitive assessment; it reflects a moral lens through which
individuals interpret inequality. People who believe that success
results from hard work and talent are more likely to view social and
economic disparities as legitimate
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_percepcion_2019}{\textbf{castillo\_percepcion\_2019?}}).
Conversely, if they see outcomes as driven by luck, social origin, or
systemic barriers, inequality is more likely to be perceived as unjust.
This distinction becomes crucial in societies with persistent structural
inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

I adopt a multidimensional perspective on meritocracy, distinguishing
between two key dimensions: effort-based and talent-based perceptions.
This distinction is essential, as it captures different pathways through
which individuals justify inequality. Effort-based meritocracy
emphasizes hard work and perseverance as the basis for social rewards,
aligning closely with cultural narratives of personal responsibility. A
talent-based meritocracy, by contrast, emphasizes intelligence and
innate abilities, which are often perceived as less malleable and more
unequally distributed. Both dimensions have been shown to correlate with
acceptance of inequality, but they may carry distinct implications for
how inequality is justified in specific domains
(\citeproc{ref-castillo_multidimensional_2023}{\textbf{castillo\_multidimensional\_2023?}}).
The relevance of this distinction is supported by recent studies, which
show that individuals respond differently to these dimensions. For
instance, perceptions that effort is rewarded in society are more
strongly associated with positive evaluations of fairness and acceptance
of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}). This may be because effort is seen as a controllable and morally
virtuous trait, whereas talent is often perceived as a natural
advantage. Consequently, effort-based meritocracy is likely more potent
in legitimizing inequality, particularly in neoliberal contexts.
